Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses

### 2021-04-13 manuscript/manuscript.Rmd
Pre-AZ edits

---
title: "Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses"
short-title: "Canary in the Coal Mine"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "April 12, 2021"
published: This is an early draft of a working paper.
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  Many countries across the globe use legal restrictions to crack down and limit civil society. These attempts to repress civil society organizations—many of which are sanctioned by law—represent a bureaucratic form of repression that could indicate more severe human rights abuses and democratic backsliding. This is especially the case for electoral democracies, which unlike autocracies do not aggressively and openly attack civic space. In this paper, we explore if crackdowns on NGOs predict broader human rights repression. We argue that this is the case because anti-NGO laws are among the most subtle means of repression and attract the least least domestic and international condemnation, as many audiences mistake this repression as regulation and are not personally affected by it. These laws also make it easier for states to prevent and deter future large-scale popular mobilization. Using original data on civil society restrictions over the last three decades, we test whether NGO crackdown is a predictor of political terror, physical violence, and violations of private civil liberties. This project has important implications: if crackdown on NGOs is indeed an early warning, it can potentially prevent further deterioration in space for civil society.
keywords: "civil society restrictions, human rights, political terror, civil liberties"
reference-section-title: References
bibliography: bibliography.bib
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Inter
editor_options: 
  chunk_output_type: console
---

@@

@@

```{r civicus-map, fig.cap="2020 CIVICUS Monitor civic space ratings"}
ggplot() +
  geom_sf(data = civicus_map_data, aes(fill = fct_rev(category)), size = 0.15, color = "black") +
  coord_sf(crs = st_crs("ESRI:54030"), datum = NA) +  # Robinson
  scale_fill_viridis_d(option = "plasma", end = 0.9, 
                       na.translate = FALSE, name = "Civic space") +
  theme_void(base_size = 9, base_family = "Inter Light") +
  theme(legend.position = "bottom",
        legend.key.size = unit(0.7, "lines"))
```

@@

Anti-civil society measures are occasionally violent, but notably, government repression of NGOs is generally non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of laws to create barriers to NGO advocacy, entry, funding [@Chaudhry:2016]. The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While in some cases this has led to the depoliticization of NGOs and avoidance of direct political activism by CSOs, states cannot soften or neutralize all forms of political activities in subtle, legal ways that they have with NGOs. Sooner or later, states need to deal with threats to their stability in a more direct, violent, confrontational manner. Administrative NGO laws and general civil society repression could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression in a state, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about a state's broader respect for civil society and the overall level of human rights repression?  Can the crackdown on NGOs act as an indicator for subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting human rights abuses, political terror, physical violence, and violations of private civil liberties. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. In some cases, such crackdown can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of civil society repression: official de jure anti-NGO restrictions that are codified in law, and general de facto levels of civil society repression. We propose and test two empirical expectations—that de jure and de factor restrictions predict future political terror, physical violence, and government encroachment on private civil liberties. Using multilevel Bayesian modeling, we find that formal de jure laws provide very little additional information when attempting to predict future repression. However, we find that accounting for more general civil society repression does boost our models' predictive power. Increasing de facto civil society repression predicts an increase in the probability of moderate political terror and results in a 3% decrease in indexes of physical violence and private civil liberties.

@@

One part of the existing literature focuses on the role of structural factors such as levels of democracy, economic development, electoral competitiveness in predisposing a state to increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

Another strand of research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. In addition to attacking groups willingness to organize against states, such preventive repression also damages states' capacity to impose costs [@Davenport:2007a, 47].

NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions or administrative crackdown can also be considered a form of preventive repression. A robust, strong civil society is widely considered as an important component on democratic transition and consolidation [@Putnam:1994]. Politically, as important civil society actors, NGOs enable citizens to voice their grievances, seek redress, and carry out lobbying and advocacy activities to further their interests. They help keep government accountable, monitor human rights violations and corruption. From an economic perspective, NGOs provide services when the state falls short. As a result of these roles, NGOs can wield real influence, which can be threatening to states. State crackdown on NGOs has become pervasive as these organizations have the ability to influence electoral politics [@DPR:2016; @Weiss:2009], aid mobilization [@MurdieBhasin:2011; @BBCM2014; @Boulding:2014], and threaten a state's economic interests [@DietrichMurdie:2017; @LebovicVoeten:2009; @Nielsen:2013].

@@

Simply noting the passage of (or counting) these laws may therefore miss the effect of their de facto implementation. Moreover, the implementation of these laws may not be limited to its legal character. The government may also engage in harassment of NGOs and activists to dissuade NGOs from operation or expressing themselves. This may include measures such as detentions, short-term incarceration, as well as beatings, threats to families and destruction of valuable property. Finally, counting official laws can also undermeasure civil society repression because governments can repress NGOs without passing or relying on formal laws. As prior research in Egypt, Ethiopia, and Russia points out, "rather than systematically enforcing all repressive legal measures-often impossible owing to capacity constraints," governments may choose to use extra-legal means to intimidate, harass or prosecute a few NGOs or activists. These cases, especially during moments or crises or contention, "highlight to the wider NGO community-and citizenry-that the government is willing and capable of using the law to repress dissent" [@Brechenmacher:2017, 94].

@@

- **Empirical expectation 1**: If a country imposes a formal anti-NGO law, then it is likely to signal future deteriorating human rights conditions.
- **Empirical expectation 2**: If a country represses civil society, then it is likely to signal future deteriorating human rights conditions.

@@

@christensen2013 collected data on formal NGO regulations in 90 countries to explore the determinants of laws that restrict NGO foreign funding laws specifically. @Chaudhry:2016 expanded this data so that it now includes official NGO laws for 162 countries between 1990 and 2013. We create indexes of the total number of laws passed in each country based on three broad categories, and we also sum each of these indexes to create a count of total legal barriers. We also create a yearly indicator variable to show if a country increased its count of anti-NGO laws (models using this outcome are included in the online appendix).

@@

Because the PTS captures broad notions of political terror with only five possible categories, we also look at three continuous measures of repression. From V-Dem, we use a physical violence index (`v2x_clphy`) that measures freedom from torture and political killings specifically, and a private civil liberties index (`v2x_clpriv`) that measures respect for property rights, labor rights, freedom of religion, freedom of movement, and other non-political human rights. V-Dem includes several other rights indicators, including one for political civil liberties that captures many of the same dimensions as the PTS, but civil society repression is included as a factor for these other indexes, making it inappropriate to use as an outcome variable. The middle panel of Figure \@ref(fig:repression) shows that the average level of these two rights indexes has improved over time.

Finally, looking at political terror, physical violence, and private civil liberties captures specific dimensions of a country's respect for human rights, but isolating these trends in repression can prove difficult. @SchnakenbergFariss:2014, @Fariss:2014, and others have argued that a country's latent and underlying human rights environment causes a state to engage in specific actions related to human rights, such as engaging in political violence, arresting activists, or restricting freedom of religion. Measuring this underlying respect for human rights is difficult, however, as it is by definition unobserved. @FarissKenwickReuning:2020 use a Bayesian measurement model to generate estimates of latent respect for physical integrity rights. We use the posterior mean of their latent variable ($\theta$) as our final measure of repression in order to see how civil society restrictions predict changes in a country's future underlying human rights environment. The right panel of Figure \@ref(fig:repression) demonstrates the global increase in average latent human rights values over time.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of NGO laws results in this turn from Level 1 to Levels 2 or 3, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of legal barriers, the overall probability of seeing Levels 4 and 5 is minuscule, and the adding new laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5, and few countries that engage in such severe repression compared to lower levels of the index. The majority of movement across PTS categories should thus occur between Levels 1–3.

@@

When holding all other factors equal, the de factor implementation of anti-civil society measures thus appear to be a stronger marginal indicator of future repression than the passing of formal de jure laws. Increasing civil society repression in countries with low levels of political terror can again potentially serve as a weak signal of more a more general increase in political violence in the future, while crackdowns on civil society predict more substantial changes in physical violence, threats to civil liberties, and latent respect for human rights more broadly.

@@

Strengthening civil society is an important precondition for democratic transition and consolidation. It keeps governments accountable and protects civilians from the worst excesses of an abusive government. When governments restrict civil society, it reduces these organizations' ability to maintain accountability and protect against abuse. Given this potential protective power, in this paper, we posited that the ongoing global crackdown on civil society could predict worsening human rights in a country interested in political and civil repression. Passing restrictive laws, imposing barriers to NGO advocacy, entry, and funding, and arresting and expelling NGO staff could pave the way for future repression.

We outline two empirical expectations: that (1) official de jure anti-NGO laws and (2) general de facto civil society restrictions would predict future repression in a country, including political terror, physical violence, and violations of private civil liberties. We find little support for the predictive power of de jure restriction. Marginal predictions show that, all else held equal, adding new anti-NGO laws increases the probability of seeing moderate levels of political terror, but does not change predicted values of physical violence or civil liberties. When comparing out-of-sample predictions to a baseline model, including de jure restrictions as predictors does not improve our models' predictive accuracy.

We do, however, find some predictive power when accounting for general civil society restrictions. As with formal laws, rising civil society repression tends to increase the probability of more moderate political terror. Additionally, increasing civil society repression is associated with a 3% decrease in indexes that measure safety from physical violence and respect for civil liberties in the future. In out-of-sample predictions, including civil society restrictions improves model accuracy by 1–2% over the baseline.

@@

>>>>

---
title: "Canary in the Coal Mine? Civil Society Repression as a Predictor of Human Rights Abuses"
short-title: "Canary in the Coal Mine"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "April 12, 2021"
published: This is an early draft of a working paper.
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  Many countries across the globe have increasingly cracked down on NGOs over the last few decades. These attempts to repress civil society organizations—many of which are sanctioned by law—represent a bureaucratic form of repression that could indicate more severe human rights abuses and democratic backsliding. This is especially the case for electoral democracies, which unlike autocracies, may not aggressively attack civic space. In this paper, we explore if crackdowns on NGOs predict broader human rights repression. Anti-NGO laws are among the most subtle means of repression and attract the least least domestic and international condemnation, as many audiences mistake this repression as regulation and are not personally affected by it. These laws may also make it easier for states to prevent and deter future large-scale popular mobilization. Using original data on civil society restrictions over the last three decades, we test whether NGO crackdown is a predictor of political terror, physical integrity rights violations, and violations of private civil liberties. We find that while formal de jure laws provide little information in predicting future repression, de facto civil society repression predicts worsening respect for physical integrity rights and civil liberties.
keywords: "civil society restrictions, human rights, political terror, civil liberties"
thanks: >-
  We would like to thank participants of the 2021 International Studies Association (ISA) convention for their comments and feedback.
reference-section-title: References
bibliography: bibliography.bib
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Inter
editor_options: 
  chunk_output_type: console
---

@@

\memSingle

@@

```{r civicus-map, fig.cap="2020 CIVICUS Monitor civic space ratings", fig.pos="h"}
ggplot() +
  geom_sf(data = civicus_map_data, aes(fill = fct_rev(category)), size = 0.15, color = "black") +
  coord_sf(crs = st_crs("ESRI:54030"), datum = NA) +  # Robinson
  scale_fill_viridis_d(option = "plasma", end = 0.9, 
                       na.translate = FALSE, name = "Civic space") +
  theme_void(base_size = 9, base_family = "Inter Light") +
  theme(legend.position = "bottom",
        legend.key.size = unit(0.7, "lines"))
```

@@

Anti-civil society measures are occasionally violent, but recently, government repression of NGOs has generally been non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of laws to create barriers to NGO advocacy, entry, and funding [@Chaudhry:2016]. The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While in some cases this has led to the depoliticization of NGOs and avoidance of direct political activism by CSOs, states cannot soften or neutralize all forms of political activities in subtle, legal ways like they have with NGOs. Sooner or later, states need to deal with threats to their stability in a more direct, violent, confrontational manner. Administrative NGO laws and general civil society repression could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression in a state, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about the state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as an indicator for subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting human rights abuses, political terror, physical violence, and violations of private civil liberties. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. In some cases, such crackdown can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of civil society repression: official de jure anti-NGO restrictions that are codified in law, and general de facto levels of civil society repression. We propose and test two empirical expectations—that de jure and de factor restrictions predict future political terror, physical integrity rights violations, and government encroachment on private civil liberties.

Using multilevel Bayesian modeling, we find that formal de jure laws provide very little additional information when attempting to predict future repression. However, we find that accounting for more general civil society repression does boost our models' predictive power. Increasing de facto civil society repression predicts an increase in the probability of moderate political terror and results in worsening violations of physical integrity rights and civil liberties.

@@

Early literature on repression focused on the role of structural factors such as levels of democracy, economic development, electoral competitiveness in predisposing a state to increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

More recent research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. In addition to attacking groups willingness to organize against states, such preventive repression also damages states' capacity to impose costs [@Davenport:2007a, 47]. NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions or administrative crackdown can also be considered a form of preventive repression [@Chaudhry:2016].

A robust, strong civil society is widely considered as an important component on democratic transition and consolidation [@Putnam:1994]. Politically, as important civil society actors, NGOs enable citizens to voice their grievances, seek redress, and carry out lobbying and advocacy activities to further their interests. They help keep government accountable, monitor human rights violations and corruption. From an economic perspective, NGOs provide services when the state falls short. As a result of these roles, NGOs can wield real influence, which can be threatening to states. State crackdown on NGOs has become pervasive as these organizations have the ability to influence electoral politics [@DPR:2016; @Weiss:2009], aid mobilization [@MurdieBhasin:2011; @BBCM2014; @Boulding:2014], and threaten a state's economic interests [@DietrichMurdie:2017; @LebovicVoeten:2009; @Nielsen:2013].

@@

Simply noting the passage of (or counting) these laws may therefore miss the effect of their de facto implementation. Moreover, the implementation of these laws may not be limited to its legal character. The government may also engage in harassment of NGOs and activists to dissuade NGOs from operation or expressing themselves. This may include measures such as detentions, short-term incarceration, as well as beatings, threats to families and destruction of valuable property.

Finally, counting official laws can also undermeasure civil society repression because governments can repress NGOs without passing or relying on formal laws. As prior research in Egypt, Ethiopia, and Russia points out, "rather than systematically enforcing all repressive legal measures-often impossible owing to capacity constraints," governments may choose to use extra-legal means to intimidate, harass or prosecute a few NGOs or activists. These cases, especially during moments or crises or contention, "highlight to the wider NGO community-and citizenry-that the government is willing and capable of using the law to repress dissent" [@Brechenmacher:2017, 94].

@@

- **Empirical expectation 1** (effect of de jure NGO repression): If a country imposes a formal anti-NGO law, then it is likely to signal future deteriorating human rights conditions.
- **Empirical expectation 2** (effect of de facto NGO repression): If a country represses civil society, then it is likely to signal future deteriorating human rights conditions.

@@

To measure de jure repression, we use data on anti-NGO laws from @Chaudhry:2016 which includes official NGO laws for 162 countries between 1990 and 2013. We create indexes of the total number of laws passed in each country based on three broad categories of laws related to barriers to entry, funding, and advocacy, and we also sum each of these indexes to create a count of total legal barriers. We also create a yearly indicator variable to show if a country increased its count of anti-NGO laws (models using this outcome are included in the online appendix).

@@

Because the PTS captures broad notions of political terror with only five possible categories, we also look at three continuous measures of repression. From V-Dem, we use a physical violence index (`v2x_clphy`) that measures freedom from torture and political killings specifically, and a private civil liberties index (`v2x_clpriv`) that measures respect for property rights, labor rights, freedom of religion, freedom of movement, and other non-political human rights. The middle panel of Figure \@ref(fig:repression) shows that the average level of these two rights indexes has improved over time.

However, looking at political terror, physical integrity rights, and private civil liberties captures specific dimensions of a country's respect for human rights, but isolating these trends in repression can prove difficult. @SchnakenbergFariss:2014, @Fariss:2014, and others have argued that a country's latent and underlying human rights environment causes a state to engage in specific actions related to human rights, such as engaging in political violence, arresting activists, or restricting freedom of religion. Measuring this underlying respect for human rights is difficult, however, as it is by definition unobserved. @FarissKenwickReuning:2020 use a Bayesian measurement model to generate estimates of latent respect for physical integrity rights. We use the posterior mean of their latent variable ($\theta$) as our final measure of repression in order to see how civil society restrictions predict changes in a country's future underlying human rights environment. The right panel of Figure \@ref(fig:repression) demonstrates the global increase in average latent human rights values over time.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of NGO laws results in this turn from Level 1 to Levels 2 or 3, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of legal barriers, the overall probability of seeing Levels 4 and 5 is minuscule, and the adding new laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5. Specifically, in addition to measuring the scope and intensity of government violence, the PTS scale also looks at the proportion of population targeted for abuse. Countries that receive a 5 are rare, as this score captures physical abuses routinely perpetrated against citizens not involved in politics, and includes cases where governments engage in "scorched earth" counterinsurgency policies or summary executions based on class or ethnic affiliation [@GibneyCornettWood:2019]. The majority of movement across PTS categories should thus occur between Levels 1–3.

@@

When holding all other factors equal, the de factor implementation of anti-civil society measures thus appear to be a stronger marginal indicator of future repression than the passing of formal de jure laws. Increasing civil society repression in countries with low levels of political terror can again potentially serve as a weak signal of more a more general increase in political violence in the future, while crackdowns on civil society predict more substantial changes in physical integrity rights, threats to civil liberties, and latent respect for human rights more broadly.

@@

Strengthening civil society is an important precondition for democratic transition and consolidation. It keeps governments accountable and protects civilians from the worst excesses of an abusive government. When governments restrict civil society, it reduces these organizations' ability to maintain accountability and protect against abuse. Given these potential knock-on effects, in this paper, we posited that the ongoing global crackdown on civil society could predict worsening human rights in a country interested in political and civil repression. Passing restrictive laws, imposing barriers to NGO advocacy, entry, and funding, and arresting and expelling NGO staff could pave the way for future repression.

We outline two empirical expectations: that (1) official de jure anti-NGO laws and (2) general de facto civil society restrictions would predict future repression in a country, including political terror, physical violence, and violations of private civil liberties. We find little support for the predictive power of de jure restriction. Marginal predictions show that, all else held equal, adding new anti-NGO laws increases the probability of seeing moderate levels of political terror, but does not change predicted values of physical integrity or civil liberties rights violations. When comparing out-of-sample predictions to a baseline model, including de jure restrictions as predictors does not improve our models' predictive accuracy.

We do, however, find some predictive power when accounting for general civil society restrictions. As with formal laws, rising civil society repression tends to increase the probability of more moderate political terror. Additionally, increasing civil society repression is associated with worsening respect for physical integrity rights and civil liberties in the future. In out-of-sample predictions, including civil society restrictions improves model accuracy by 1–2% over the baseline.

@@

\memSingleSmall
